---
title: "Project"
author: "Andrei Graham, Arlette Hall, Tyler Sousa"
date: "2026-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

1. Clean Data
    - clean names of variables
    - remove any NAs 
    - rename lengthier variable names
    - translate coordinates from European to longitude, latitude
    - save cleaned data as own .csv file
    
2. Data Exploration
    - Arlette: Are there any trends for number of redlisted species and landscape/plot types?
      - What could this mean?
    - Tyler: Are there any trends for total biomass and landscape/plot types?
      - What could this mean?
    - Andrei: Are there any hidden trends for total biomass and weather (temperature, humidity, precipitation)?
      - What could this mean?

3. Shiny App
    - display total biomass or number of redlisted species per plot
      - can change landscape type
      - can change plot type

## Load the Libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(naniar)
library(sf)
library(ggmap)
library(paletteer)
library(shiny)
library(shinydashboard)
```

## Load in Data
```{r}
insects <- read_csv2("data/insect_raw/Data_Uhler_2021_main.csv") %>%
  clean_names()
```

## Check Data Structure
```{r}
insects %>%
  glimpse()
```

## Clean Data - Arlette
```{r}
#clean possible hidden NAs from character variables
insects <- insects %>%
  replace_with_na_all(condition = ~.x %in% c(common_na_strings))
```

```{r}
#check numeric variables for hidden NAs
insects %>%
  select(-c(plotid,
             quadrant_id,
             landscape_type,
             landscape_type_2,
             plot_type,
             plot_type_2,
             landscape_plot,
             date_start,
             date_end)) %>%
  summary()

#Per insect_README.txt, values of 0 seen here are not NAs.
#No hidden NAs.
```

```{r}
#view NA summary
insects %>%
  miss_var_summary
```

```{r}
#rename average_temperature to avg_temp
insects <- insects %>%
  rename(avg_temp = average_temperature) %>%
#rename average_humidity to avg_humidity
  rename(avg_humidity = average_humidity) %>%
#rename long_term_mean_annual_precipitation to lt_mean_annual_precip
  rename(lt_mean_annual_precip = long_term_mean_annual_precipitation) %>%
#rename long_term_mean_annual_average_near_surface_temperature to lt_mean_annual_avg_ns_temp
  rename(lt_mean_annual_avg_ns_temp = long_term_mean_annual_average_near_surface_temperature)
```

```{r}
#translate coordinates from ETRS89/UTM zone 32N to WGS84, EPSG:4326 (longitude, latitude)

#give each sample a unique ID #
insects <- insects %>%
  mutate(sample_id = row_number())
#copy point_x, point_y, and sample_id to new temp table
temp_coord_table <- insects %>%
  select(point_x, point_y, sample_id)
#convert and save coordinates
temp_coord_table <- temp_coord_table %>%
  st_as_sf(coords = c("point_x", "point_y"), crs = 25832) %>%
  st_transform(crs = 4326) %>%
  mutate(lat = st_coordinates(.)[,1], long = st_coordinates(.)[,2] - 30)
#join temp table to insects table using full_join()
insects <- full_join(insects, temp_coord_table, by="sample_id")
```

```{r}
#check structure of new columns
insects %>%
  select(sample_id, long, lat, geometry) %>%
  glimpse()

#All good, nothing left to correct.
#No date splitting in main cleaned data to avoid column clutter (start_month, end_month, etc.)
```

```{r}
write_csv(insects, "data/insects_main.csv")
```

## Data Exploration - Arlette

My Question: How does the number of red-listed species compare across landscape & plot type?

```{r}
#load cleaned data as own variable (to not alter others' work)
insects_ah <- read.csv("data/insects_main.csv")
```

```{r}
#check data structure
insects_ah %>%
  glimpse()
```

landscape_type & plot_type:
```{r}
#Variable Definitions:
# RL_Richness = No. of species on red-lists for Bavaria/Germany per plot & time frame.
# Landscape_Type = Landscape category of quadrant (agricultural, semi-natural, or urban).
# Plot_Type = Habitat category of plot (arable field, forest, meadow, natural, or settlement).
# Landscape (Quadrant) > Plot

#check landscape_type vs rl_richness
insects_ah %>%
  group_by(landscape_type) %>%
  select(landscape_type, rl_richness) %>%
  na.omit() %>%
  summarize(total_rl_richness = sum(rl_richness), n = n()) %>%
  arrange(desc(total_rl_richness))

#Sampling is even across landscape types
#Semi-natural >> Urban > Agricultural (pesticides?)
```

```{r}
#check plot_type vs rl_richness
insects_ah %>%
  group_by(plot_type) %>%
  select(plot_type, rl_richness) %>%
  na.omit() %>%
  summarize(total_rl_richness = sum(rl_richness), n = n()) %>%
  arrange(desc(total_rl_richness))

#Sampling uneven across plot types. Forests may have just been more prevalent.
#Arable fields sampling on par with meadow & greater than settlement but have much less redlist species.
#Forest >>> Meadow > Settlement >> Arable Field (pesticides?)
```

```{r}
#cluster column chart of landscape_type vs rl_richness by plot_type
insects_ah %>%
  group_by(landscape_type, plot_type) %>%
  select(landscape_type, plot_type, rl_richness) %>%
  na.omit() %>%
  summarize(total_rl_richness = sum(rl_richness),
            .groups='keep') %>%
  ggplot(mapping=aes(x = reorder(plot_type, total_rl_richness),
                     y = total_rl_richness,
                     fill = landscape_type))+
  geom_col(position = "dodge", color = "black", na.rm=T)+
  coord_flip()+
  labs(x = "Plot Type", y = NULL,
       title = "No. of Redlist Species Found by Plot Type & Landscape",
       fill = "Landscape Type")+
  theme_linedraw()
```

```{r}
#boxplot of landscape_type vs rl_richness by plot_type
insects_ah %>%
  group_by(landscape_type, plot_type) %>%
  select(landscape_type, plot_type, rl_richness) %>%
  na.omit() %>%
  ggplot(mapping=aes(x = reorder(plot_type, rl_richness),
                     y = rl_richness,
                     fill = landscape_type))+
  geom_boxplot(na.rm = T)+
  labs(x = "Plot Type", y = NULL,
       title = "No. of Redlist Species Found by Plot Type & Lanscape",
       fill = "Landscape Type")+
  theme_linedraw()
```

Build a Map:
```{r}
#register
register_stadiamaps("1ad11010-4c59-42b0-a160-0dab4e61577f", write = FALSE)
```

```{r}
#get mins and maxes for longitude & latitude
insects_ah %>%
  select(c(lat, long, rl_richness)) %>%
  filter(lat != "NA" & long != "NA" & rl_richness != "NA") %>% #omit NAs
  summary()
```

```{r}
#set latitude and longitude ranges
lat <- c(46.12, 48.76)
long <- c(10.59, 13.41)
#make bbox with 3% (f=0.03) buffer
bbox <- make_bbox(long, lat, f=0.03)
```

```{r}
#load map
map <- get_stadiamap(bbox, maptype = "stamen_terrain", zoom = 7)
```

```{r}
#plot map by landscape type
ggmap(map)+
  geom_point(data=insects_ah %>%
               filter(rl_richness > 0),
             aes(long, lat, shape = landscape_type), color="black", size=2, alpha=0.3)+
  labs(x="Longitude", y="Latitude", title="Locations of Redlist Species by Landscape Type")
```

```{r}
#plot map by plot type
ggmap(map)+
  geom_point(data=insects_ah %>%
               filter(rl_richness > 0),
             aes(long, lat, shape = plot_type), color="black", size=2, alpha=0.3)+
  labs(x="Longitude", y="Latitude", title="Locations of Redlist Species by Plot Type")
```

Conclusion: The number of redlisted species is lowest in agricultural landscape & arable fields. Possible reasons are the use of pesticides or lack of edible plants (food grown for humans is not always edible to native insects).

## Data Exploration - Tyler

My Question: How does the biomass collected (g) compare across landscape & plot type?

```{r}
#load cleaned data as own variable (to not alter others' work)
insects_ts <- read_csv("data/insects_main.csv")
```

```{r}
#Variable Definitions:
# Biomass = Insect biomass in g.
# Landscape_Type = Landscape category of quadrant (agricultural, semi-natural, or urban).
# Plot_Type = Habitat category of plot (arable field, forest, meadow, natural, or settlement).

#Check landscape_type vs biomass
insects_ts %>%
  group_by(landscape_type) %>%
  select(landscape_type, biomass) %>%
  na.omit() %>%
  summarize(total_biomass = sum(biomass), n = n()) %>%
  arrange(desc(total_biomass))

#Sampling is not even across landscape types.
```

```{r}
#Check plot_type vs biomass
insects_ts %>%
  group_by(plot_type) %>%
  select(plot_type, biomass) %>%
  na.omit() %>%
  summarize(total_biomass = sum(biomass), n = n()) %>%
  arrange(desc(total_biomass))

#Sampling is not even across plot types.
```

```{r}
#boxplot of landscape_type vs biomass by plot_type
insects_ts %>%
  group_by(landscape_type, plot_type) %>%
  select(landscape_type, plot_type, biomass) %>%
  na.omit() %>%
  ggplot(aes(x = reorder(landscape_type, biomass), y = biomass, fill = plot_type))+
  geom_boxplot()+
  labs(title = "Range of Biomass Found by Landscape and Plot Type",
       x="Landscape Type",
       y = "Biomass (g)",
       fill = "Plot Type")+
  theme_linedraw()
```

```{r}
#boxplot of plot_type vs biomass by landscape_type
insects_ts %>%
  group_by(landscape_type, plot_type) %>%
  select(landscape_type, plot_type, biomass) %>%
  na.omit() %>%
  ggplot(aes(x = reorder(plot_type, biomass), y = biomass, fill = landscape_type))+
  geom_boxplot()+
  labs(title = "Range of Biomass Found by Plot and Landscape Type",
       x="Plot Type",
       y = "Biomass (g)",
       fill = "Landscape Type")+
  theme_linedraw()
```

```{r}
#Check landscape_type vs biomass
insects_ts %>%
  group_by(plot_type, landscape_type) %>%
  select(landscape_type,plot_type, biomass) %>%
  na.omit() %>%
  summarize(total_biomass = sum(biomass), n = n(),
            .groups = 'keep') %>%
  arrange(desc(n))

#Sampling is not even across plot/landscape types. Differences in the total biomass below are likely influenced sampling differences. 
```

```{r}
#graph of the total biomass collected in every landscape/plot type combina
insects_ts %>% 
  group_by(plot_type, landscape_type) %>% 
  select(landscape_type, plot_type, biomass) %>%
  na.omit() %>%
  summarize(tot_biomass=sum(biomass),
            .groups = 'keep') %>% 
  mutate(land_plot=paste(plot_type, landscape_type, sep = "/")) %>%
  ggplot(aes(x=tot_biomass, y=reorder(land_plot, tot_biomass), fill=land_plot))+
  geom_col()+
  labs(title="Total Insect Biomass per Land/Plot Type",
       y=NULL,
       x="Total Biomass (g)")+
  theme_linedraw()+
  theme(legend.position = "none")
```

```{r}
#boxplot of plot_type vs biomass by landscape_type
insects_ts %>%
  filter(excluded!=1) %>% 
  group_by(plotid) %>% 
  summarize(biomass=sum(biomass), landscape_type=landscape_type[1], plot_type=plot_type[1]) %>% 
  select(landscape_type, plot_type, biomass) %>%
  na.omit() %>%
  ggplot(aes(x = reorder(plot_type, biomass), y = biomass, fill = landscape_type))+
  geom_boxplot()+
  labs(title = "Range of Total Plot Biomass by Plot Type and Landscape Type",
       x="Plot Type",
       y = "Total Plot Biomass (g)",
       fill = "Landscape Type")+
  theme_linedraw()
```

Conclusion: The total biomass is lowest in urban landscapes & settlements. This points to the trend seen in redlisted species is due to a lack of edible plants in arable fields (food grown for humans is not always edible to native insects).


## Data Exploration - Andrei

My Question: Are there hidden factors from weather (temperatur, humidity, precipitation) influencing total biomass trends?

```{r}
#load cleaned data as own variable (to not alter others' work)
insects_ag <- read.csv("data/insects_main.csv")
```

```{r}
#check data structure
glimpse(insects_ag)
```

```{r}
#map
insects_ag %>%
  filter(biomass != "NA")%>%
  select(lat, long) %>% 
  summary()
```

```{r}
register_stadiamaps("824e1e2a-26b0-4511-874e-b93a502f5dce", write = FALSE)
```

```{r}
lat<- c(46.12, 48.76)
long<- c(10.59, 13.41)
bbox<- make_bbox(long, lat, f=0.05)
```

```{r}
#map of where the data came from (hopefully)
map1<- get_stadiamap(bbox, maptype = "stamen_terrain", zoom = 7)
```

```{r}
ggmap(map1)+
  geom_point(data=insects_ag %>%
               filter(long != "NA" & lat != "NA" & biomass!="NA"), aes(long, lat), size=0.8)+
  labs(x="Longitude", y="Latitude", title="Sites Where Biomass was Found")
```

```{r}
#average humidity and biomass
insects_ag%>%
  filter(biomass != "NA" & avg_humidity != "NA")%>%
  group_by(avg_humidity)%>%
  summarise(total_biomass=sum(biomass), n=n())%>%
  select(avg_humidity, total_biomass, n)%>%
  arrange(desc(n))
```

```{r}
#plot average humidity and biomass
insects_ag%>%
  filter(biomass != "NA" & avg_humidity != "NA")%>%
  group_by(avg_humidity)%>%
  summarise(total_biomass=sum(biomass))%>%
  ggplot(aes(x = avg_humidity, y = total_biomass))+
  geom_point(na.rm=TRUE)+
  geom_smooth(method=lm, se=T, na.rm=T)+
  labs(x = "Average Humidity (%)",
       y = "Total Biomass (g)",
       title = "Average Humidity vs Total Biomass")+
  theme_linedraw()

#no strong trend found
```

```{r}
#average temperature and biomass
insects_ag%>%
  filter(biomass != "NA" & avg_temp != "NA")%>%
  group_by(avg_temp)%>%
  summarise(total_biomass=sum(biomass), n=n())%>%
  select(avg_temp, total_biomass, n)%>%
  arrange(desc(n))
```

```{r}
#plot average temperature and biomass
insects_ag%>%
  filter(biomass != "NA" & avg_temp != "NA")%>%
  group_by(avg_temp)%>%
  summarise(total_biomass=sum(biomass))%>%
  ggplot(aes(x = avg_temp, y = total_biomass))+
  geom_point(na.rm=TRUE)+
  geom_smooth(method=lm, se=T, na.rm=T)+
  labs(x = "Average Temperature (C)",
       y = "Total Biomass (g)",
        title = "Average Temp vs Total Biomass")+
  theme_linedraw()

#no strong trend found
```

```{r}
#longterm mean annual precipitation and biomass
insects_ag%>%
  filter(lt_mean_annual_precip != "NA" & biomass != "NA")%>%
  group_by(lt_mean_annual_precip)%>%
  summarise(total_biomass=sum(biomass), n=n())%>%
  select(lt_mean_annual_precip, total_biomass, n)%>%
  arrange(desc(n))
```

```{r}
#plot longterm mean annual precipitation and biomass
insects_ag%>%
  filter(lt_mean_annual_precip != "NA" & biomass != "NA")%>%
  group_by(lt_mean_annual_precip)%>%
  summarise(total_biomass=sum(biomass))%>%
  ggplot(aes(x = lt_mean_annual_precip, y = total_biomass))+
  geom_point(na.rm=TRUE)+
  geom_smooth(method=lm, se=T, na.rm=T)+
  labs(x = "Longterm Mean Annual Precipitation (mm)",
       y = "Total Biomass (g)",
        title = "Longterm Mean Annual Precipitation vs Total Biomass")+
  theme_linedraw()

#no strong trend found
```

```{r}
#longterm mean annual average-near surface temperature and biomass
insects_ag%>%
  filter(lt_mean_annual_avg_ns_temp != "NA" & biomass != "NA")%>%
  group_by(lt_mean_annual_avg_ns_temp)%>%
  summarise(total_biomass=sum(biomass), n=n())%>%
  select(lt_mean_annual_avg_ns_temp, total_biomass, n)%>%
  arrange(desc(n))
```

```{r}
#plot longterm mean annual average-near surface temperature and biomass
insects_ag%>%
  filter(lt_mean_annual_avg_ns_temp != "NA" & biomass != "NA")%>%
  group_by(lt_mean_annual_avg_ns_temp)%>%
  summarise(total_biomass=sum(biomass), n=n())%>%
  ggplot(aes(x = lt_mean_annual_avg_ns_temp, y = total_biomass))+
  geom_point(na.rm=TRUE)+
  geom_smooth(method=lm, se=T, na.rm=T)+
  labs(x = "Longterm Mean Annual Average-Near Surface Temperature (C)",
       y = "Total Biomass (g)",
        title = "Longterm Mean Annual Average-Near Surface Temperature vs Total Biomass")+
  theme_linedraw()

#no strong trend found
```

Conclusion: There are no hidden causal factors in the weather (temperature, humidity, precipitation) on trends seen in total biomass.

## Shiny App - Tyler, Arlette, Andrei
```{r, eval=F, echo=T}
ui <- dashboardPage(
  
  dashboardHeader(title="Take A Closer Look!"),
  
  dashboardSidebar(disable = T),
  
  dashboardBody(
    
    h3("Choose Landscape Type and Plot Type:"),
    
    radioButtons("lnd_type", "Landscape Type:",
               choices = c("All", "Agricultural", "Semi-natural", "Urban"),
               selected = "All", inline = TRUE),
    
    radioButtons("plt_type", "Plot Type:",
               choices = c("All", "settlement", "meadow", "arable field", "forest"),
               selected = "All", inline = TRUE),
    
    h3("Select Plot:"),
    
    selectInput("x",
              "Select Plot",
              choices=NULL),
    
    #plotOutput("plot", width="500px", height="400px")
    h3("Plot Info:"),
    
    h4("Total Biomass Collected In Grams:"),
    textOutput("biomass"),
    
    h4("Total Number of Species on the Red List:"),
    textOutput("rl")
  )
)

server <- function(input, output, session) {
  
  observe({
    updated_choices <- insects_ts
    
    if (input$lnd_type != "All") {
      updated_choices <- updated_choices %>% filter(landscape_type == input$lnd_type)
    }
    
    if (input$plt_type != "All") {
      updated_choices <- updated_choices %>% filter(plot_type == input$plt_type)
    }
    
  ids <- sort(unique(updated_choices$plotid))
  
  updateSelectInput(session, "x", choices = ids)
  
  })
  
  output$biomass <- renderText({
    req(input$x) 
    insects_ts %>% 
      filter(plotid == input$x, excluded!=1) %>% 
      group_by(plotid) %>% 
      summarize(biomass_tot=sum(biomass)) %>% 
      pull(biomass_tot) %>% 
      as.character()
  })
  
  output$rl <- renderText({
    req(input$x) 
    insects_ts %>% 
      filter(plotid == input$x, excluded!=1) %>% 
      group_by(plotid) %>% 
      summarize(rl_richness_tot=sum(rl_richness, na.rm = T)) %>% 
      pull(rl_richness_tot) %>% 
      as.character()
  })
  
}

shinyApp(ui, server)
```

